---
output: 
  pdf_document:
    keep_tex: true
    fig_caption: true
    latex_engine: xelatex
    number_sections: true
    toc: true
    toc_depth: 4
title: "SURVIVAL ANALYSIS - TCGA PRAD CANCER"
author: 
- Kelvin Ofori-Minta
- University of Texas at El Paso (UTEP) 
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontsize: 11pt
spacing: single
header-includes:
- \usepackage{amsmath}
- \usepackage{amssymb, bm}
- \usepackage{amsfonts}
- \usepackage{amsthm}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhf{}
- \rhead{Collaborative Research}
- \lhead{Thyroid Cancer}
- \cfoot{\thepage}
- \usepackage{algorithm}
- \usepackage[noend]{algpseudocode}
---

\section{Loading and Cleaning Data}
```{r, warning=FALSE, message=FALSE}
data <- read.csv("thyroidcancer.csv", header = T, na.strings = "NA")

data[data==""]<-NA #replace all empty cells with na

# write.csv(data,"C://Users//Kelvin//Desktop//Spring 2022//research with Dr. Leung//survival//selected_columns.csv")
```

\subsection{Inspecting dataframe for missing values}
```{r, warning=FALSE, message=FALSE}
require(inspectdf)
show_plot(inspect_na(data))
missing = inspect_na(data)
missing[ , 3] = round(missing[ ,3], 2)
names(missing) = c("variable", "count", "proportion")
require(kableExtra)
kable(missing)
```

\newpage
\subsubsection{Inspect distribution of variables}
```{r, warning=FALSE, message=FALSE}
explorecolumns_thyroid=c("AGE","SEX","PERSON_NEOPLASM_CANCER_STATUS","ETHNICITY", 
                           "RACE","RADIATION_THERAPY", "PFS_STATUS") 
dat=data[,explorecolumns_thyroid]

colnames(dat)[colnames(dat)=="PERSON_NEOPLASM_CANCER_STATUS"]<-"neoplasm_status"

dat$PFS_STATUS<-as.integer(ifelse(dat$PFS_STATUS=="0:CENSORED",0,1)) 
dat$RADIATION_THERAPY = as.integer(ifelse(dat$RADIATION_THERAPY=="No",0,1))
dat$SEX = as.integer(ifelse(dat$SEX=="Female",0,1))
dat$neoplasm_status=as.integer(ifelse(dat$neoplasm_status=="Tumor Free",0,1))

suppressPackageStartupMessages(library(tidyverse))
dat%>%
  pivot_longer(cols=c(PFS_STATUS,RADIATION_THERAPY,AGE, SEX,neoplasm_status),
               names_to ="key", values_to = "value", drop_na(dat)) %>%
  ggplot(aes(value)) +
  geom_histogram(bins = 20) +
  facet_wrap(~key, scales='free_x')
```


\newpage
\subsubsection{Re-coding variables}
```{r, warning=FALSE}
data=data[ ,-c(28)] #remove weight,it has empty cells


data$RADIATION_THERAPY = factor(data$RADIATION_THERAPY, levels = c("No","Yes"),
                                labels =c("No", "Yes"))

data$SEX = factor(data$SEX, levels=c("Female","Male"), labels=c("Female","Male"))

data$AJCC_PATHOLOGIC_TUMOR_STAGE=factor(data$AJCC_PATHOLOGIC_TUMOR_STAGE,levels = c("STAGE I","STAGE II","STAGE III","STAGE IV","STAGE IVA","STAGE IVC"),labels=c("STAGE I","STAGE II","STAGE III","STAGE IV","STAGE IVA","STAGE IVC"))

data$AJCC_STAGING_EDITION = factor(data$AJCC_STAGING_EDITION,
                                   levels = c("4TH","5TH","6TH","7TH"),
                                   labels = c("4TH","5TH","6TH","7TH"))
data$ETHNICITY=factor(data$ETHNICITY, 
                      levels=c("Hispanic Or Latino", "Not Hispanic Or Latino"),
                      labels =c("Hispanic Or Latino", "Not Hispanic Or Latino"))

data$PFS_STATUS<-as.integer(ifelse(data$PFS_STATUS=="0:CENSORED",0,1)) 
```



\newpage
\section{KM Curve - PF Survival of patients with Radiation Therapy}
```{r, warning=FALSE, message=FALSE}
library("survival")
library("survminer")
ndata<-data
fit1<-survfit(Surv(ndata$PFS_MONTHS, ndata$PFS_STATUS==1)~ndata$RADIATION_THERAPY
              ,data=ndata)
print(fit1)
summary(fit1)$table
```
\newpage
```{r, warning=FALSE, message=FALSE}
ggsurvplot(fit1,
          #legend.labs=c("tumor_free", "with_tumor"),
          pval = TRUE, conf.int = F,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#E7B800", "#2E9FDF"))
```

\newpage
\section{PF Survival of Neoplasm Tumor Patients Exposed to Radiation Therapy}
```{r}
tumor=ndata[ndata$PERSON_NEOPLASM_CANCER_STATUS=="With Tumor",]

fit2<-survfit(Surv(tumor$PFS_MONTHS, tumor$PFS_STATUS==1)~tumor$RADIATION_THERAPY
              ,data=tumor)

print(fit2)
summary(fit2)$table
```

```{r, warning=FALSE, message=FALSE}
ggsurvplot(fit2,
          pval = TRUE, conf.int = F,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#E7B800", "#2E9FDF"))


table(data$PERSON_NEOPLASM_CANCER_STATUS)
table(tumor$PFS_STATUS)
table(tumor$RADIATION_THERAPY)
```


\newpage
\section{Logrank Test}
```{r}
logrank <- survdiff(Surv(tumor$PFS_MONTHS, tumor$PFS_STATUS==1)~tumor$RADIATION_THERAPY
              ,data=tumor)
logrank
```


\newpage
\section{Cox Proportional Hazard Model with Neoplasm Tumor Data}
```{r, warning=FALSE, message=FALSE}
fitph<-coxph(Surv(PFS_MONTHS,PFS_STATUS==1) ~ RADIATION_THERAPY +
                AGE + SEX + RACE + AJCC_PATHOLOGIC_TUMOR_STAGE + DAYS_TO_BIRTH
                + DAYS_LAST_FOLLOWUP + AJCC_PATHOLOGIC_TUMOR_STAGE +
               AJCC_STAGING_EDITION, 
             data=tumor)
#summary(fitph)
```


\newpage
\section{Formating Cox Regression Results}
```{r}
broom::tidy(fitph , 
            exp=TRUE) %>%
  kable()
```

```{r}
fitph %>%
  gtsummary::tbl_regression(exp=TRUE)
```






























