---
output: 
  pdf_document:
    keep_tex: true
    fig_caption: true
    latex_engine: xelatex
    number_sections: true
    toc: true
    toc_depth: 4
title: "SURVIVAL ANALYSIS - TCGA PRAD CANCER"
author: 
- Kelvin Ofori-Minta
- University of Texas at El Paso (UTEP) 
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontsize: 11pt
spacing: single
header-includes:
- \usepackage{amsmath}
- \usepackage{amssymb, bm}
- \usepackage{amsfonts}
- \usepackage{amsthm}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhf{}
- \rhead{Collaborative Research}
- \lhead{TCGA - Clinical Data}
- \cfoot{\thepage}
- \usepackage{algorithm}
- \usepackage[noend]{algpseudocode}
---

\newpage
\section{Loading and Cleaning Data}
```{r, warning=FALSE, message=FALSE}
s <- read.csv("tcga_prad.csv", header = T, stringsAsFactors = F, na.strings = "NA")
#selecting columns of interest
data<-s[, c(4,7,8,12,13,19,20,21,23,27:32,44:47,50,53:55,58,62,40:42)]
# write.csv(data,"C://Users//Kelvin//Desktop//Spring 2022//research with Dr. Leung//survival//selected_columns.csv")
```

\subsection{Inspecting dataframe for missing values }
```{r, warning=FALSE, message=FALSE}
require(inspectdf)
show_plot(inspect_na(data))
missing = inspect_na(data)
missing[ , 3] = round(missing[ ,3], 2)
names(missing) = c("variable", "count", "proportion")
require(kableExtra)
# missing<-as.matrix.data.frame(missing)
kable(missing)
# as.data.frame.matrix(missing)
# kable(as.da(missing))
```

\subsubsection{Rename long variables_kkkiiiiiii}
```{r}
colnames(data)[colnames(data)=="Last.Communication.Contact.from.Initial.Pathologic.Diagnosis.Date"]<-"last_contact_from_ initial_diagnosis"

colnames(data)[colnames(data)=="Neoadjuvant.Therapy.Type.Administered.Prior.To.Resection.Text"]<-"Neodjuvant_Therapy_type"

colnames(data)[colnames(data) =="New.Neoplasm.Event.Post.Initial.Therapy.Indicator"] <- "New_Tumor_event"

colnames(data)[colnames(data) =="Person.Neoplasm.Cancer.Status"] <- "Neoplasm_status"

colnames(data)[colnames(data)=="Progress.Free.Survival..Months."] <- "progression_time"

colnames(data)[colnames(data)=="Progression.Free.Status"] <- "progression_status"


newdata<-data
"TMB-H means that the tumor has a high number of mutations. Doctors have found that
certain immunotherapy drugs are more likely to work
against TMB-H cancers. This is because the immune
system may be able to find and attack cancer cells with high
TMB more easily."


"Person neoplasm status...... You are correct, IMO:  tumor free does not mean normal, but rather whether (or not) the tumor (neoplasm) continues to be present.  It is a statement about the progression (or not) of the original disease; whereas normal is a statement that there was no disease to begin with."
```


\newpage
\subsubsection{Re-coding variables}
```{r, warning=FALSE, message=FALSE}
#newdata$Neodjuvant_Therapy_type<-as.integer(ifelse(newdata$Neodjuvant_Therapy_type=="No",1,2))

newdata$Neodjuvant_Therapy_type <- factor(newdata$Neodjuvant_Therapy_type,
                                    levels=c("No","Yes"),
                                    labels=c("No","Yes"))


# newdata$In.PanCan.Pathway.Analysis<-as.numeric(ifelse(newdata$In.PanCan.Pathway.Analysis=="No",1, 2))

newdata$In.PanCan.Pathway.Analysis<-factor(newdata$In.PanCan.Pathway.Analysis,
                                           levels=c("No","Yes"),
                                           labels=c("No","Yes"))

# newdata$Prior.Diagnosis<-as.integer(ifelse(newdata$Prior.Diagnosis=="No", 
#                                    ifelse(newdata$Prior.Diagnosis=="Yes",2,3)))

newdata$Prior.Diagnosis<-factor(newdata$Prior.Diagnosis, 
                                levels=c("No","Yes","Yes, History Of Synchronous And Or Bilateral Malignancy"),
                                labels=c("No","Yes", "yes_history"))
  
# newdata$tissue_source.site<-as.integer(ifelse(newdata$tissue_source.site=="university",1,
#                              ifelse(newdata$tissue_source.site=="Biotech & Pharma",2,
#                              ifelse(newdata$tissue_source.site=="Hospital",3,4))))


newdata$tissue_source.site<-factor(newdata$tissue_source.site,
                                   levels = c("university","Biotech & Pharma","Hospital","Research center"),
                                   labels=c("university","biotech_pharma","hospital","research_centers"))

# newdata$New_Tumor_event<-as.integer(ifelse(newdata$New_Tumor_event=="No",1,2))

newdata$New_Tumor_event <- factor(newdata$New_Tumor_event,
                                    levels=c("No","Yes"),
                                    labels=c("No","Yes"))

# newdata$Radiation.Therapy<-as.integer(ifelse(newdata$Radiation.Therapy=="No",1,2))


newdata$Radiation.Therapy <- factor(newdata$Radiation.Therapy,
                                    levels=c("No","Yes"),
                                    labels=c("No","Yes"))

newdata$Neoplasm_status <- factor(newdata$Neoplasm_status,
                                    levels=c("Tumor Free","With Tumor"),
                                    labels=c("tumor_free","with_tumor"))

#newdata$Neoplasm_status<-as.integer(ifelse(newdata$Neoplasm_status=="Tumor Free",1,2))

newdata$Race.Category<-factor(newdata$Race.Category,
                              levels = c("Asian", "Black or African American ", "White"),
                              labels = c("asian", "b/a_american","white"))

# newdata$Race.Category<-as.integer(ifelse(newdata$Race.Category=="Asian",1,
#                                          ifelse(newdata$Race.Category=="White",2,3)))


newdata$progression_status<-as.integer(ifelse(newdata$progression_status=="0:CENSORED",1,2))

str(newdata)
```

\newpage
\section{KM Curve- Default survival probability with Neoplasm status}
```{r, warning=FALSE, message=FALSE}
library("survival")
library("survminer")

fit1a <- survfit(Surv(newdata$progression_time, newdata$progression_status) ~ newdata$Neoplasm_status, 
               data = newdata)
print(fit1a)

summary(fit1a)$table
```
```{r}
ggsurvplot(fit1a,
          #legend.labs=c("tumor_free", "with_tumor"),
          pval = TRUE, conf.int = F,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#E7B800", "#2E9FDF"))

#1 - censored & 2- progression
#1 - tumor_free & 2 with tumor .....neoplasm status 
#1 - NO & 2-YES .....TREATMENT CODE
```




\newpage
\section{KM Curve- survival probability:Censored cases of Neoplasm Status}
```{r, warning=FALSE, message=FALSE}
library("survival")
library("survminer")

fit1b <- survfit(Surv(newdata$progression_time, newdata$progression_status==1) ~ newdata$Neoplasm_status, 
               data = newdata)
print(fit1b)

summary(fit1b)$table
```
```{r}
ggsurvplot(fit1b,
          #legend.labs=c("tumor_free", "with_tumor"),
          pval = TRUE, conf.int = F,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#E7B800", "#2E9FDF"))

#1 - censored & 2- progression
#1 - tumor_free & 2 with tumor .....neoplasm status 
#1 - NO & 2-YES .....TREATMENT CODE
```


\newpage
\section{KM Curve - survival probability: uncensored cases}
```{r, warning=FALSE, message=FALSE}
library("survival")
library("survminer")

fit1c <- survfit(Surv(newdata$progression_time, newdata$progression_status==2) ~ newdata$Neoplasm_status, 
               data = newdata)
print(fit1c)

summary(fit1c)$table

#equivalent to Default method
```

```{r}
ggsurvplot(fit1c,
          #legend.labs=c("tumor_free", "with_tumor"),
          pval = TRUE, conf.int = F,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#E7B800", "#2E9FDF"))

#1 - censored & 2- progression
#1 - tumor_free & 2 with tumor .....neoplasm status 
#1 - NO & 2-YES .....TREATMENT CODE
```


\newpage
\section{KM Curve- survival probability after survey days}
```{r, warning=FALSE, message=FALSE}
library("survival")
library("survminer")

fit1d <- survfit(Surv(newdata$progression_time, newdata$progression_status==2) ~ 1, 
               data = newdata)
print(fit1d)

summary(fit1d)$table
```

```{r}
ggsurvplot(fit1d,
          pval = TRUE, conf.int = T,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#E7B800", "#2E9FDF"))

#1 - censored & 2- progression
#1 - tumor_free & 2 with tumor .....neoplasm status 
#1 - NO & 2-YES .....TREATMENT CODE
```
PFS survival probability after a certain number of days


\newpage
\section{KM Curve- survival probability of neoadjuvant therapy type uncensored cases}
```{r, warning=FALSE, message=FALSE}
library("survival")
library("survminer")

fit1e<-survfit(Surv(newdata$progression_time, newdata$progression_status==2) ~ newdata$Neodjuvant_Therapy_type, 
               data = newdata)
print(fit1e)

summary(fit1e)$table
```

```{r}
ggsurvplot(fit1e,
          # legend.labs=c("No", "Yes"),
          pval = TRUE, conf.int = F,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#E7B800", "#2E9FDF"))

#1 - censored & 2- progression
#1 - tumor_free & 2 with tumor .....neoplasm status 
#1 - NO & 2-YES .....TREATMENT CODE
```

\newpage
\section{KM Curve- survival probability with New Tumor Event for uncensored cases}
```{r, warning=FALSE, message=FALSE}
library("survival")
library("survminer")

fit1f <- survfit(Surv(newdata$progression_time, newdata$progression_status==2) ~ newdata$New_Tumor_event, 
               data = newdata)
print(fit1f)

summary(fit1f)$table
```

```{r}
ggsurvplot(fit1f,
           #legend.labs=c("no", "yes"),
          pval = TRUE, conf.int = F,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#E7B800", "#2E9FDF"))

#1 - censored & 2- progression
#1 - tumor_free & 2 with tumor .....neoplasm status 
#1 - NO & 2-YES .....TREATMENT CODE
```


\newpage
\section{KM Curve- survival probability: Radiation therapy uncensored}
```{r, warning=FALSE, message=FALSE}
library("survival")
library("survminer")

fit2a <- survfit(Surv(newdata$progression_time, newdata$progression_status==2)
                 ~newdata$Radiation.Therapy
                   , data = newdata)
print(fit2a)

summary(fit2a)$table
```

```{r}
ggsurvplot(fit2a,
          pval = TRUE, conf.int = F, 
          #legend.labs=c("no_therapy", "therapy"),
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#E7B800", "#2E9FDF"))

#1 - censored & 2- progression
#1 - tumor_free & 2 with tumor .....neoplasm status 
#1 - NO & 2-YES .....TREATMENT CODE
#1 - no & 2-yes......radiation therapy
```

\newpage
\section{KM Curve- survival probability: Radiation therapy censored cases}
```{r, warning=FALSE, message=FALSE}
library("survival")
library("survminer")

fit2b<- survfit(Surv(newdata$progression_time, newdata$progression_status==1) ~
                   newdata$Radiation.Therapy, data = newdata)
print(fit2b)

summary(fit2b)$table
```

```{r}
ggsurvplot(fit2b,
          pval = TRUE, conf.int = F,
          #legend.labs=c("no_therapy", "therapy"),
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#E7B800", "#2E9FDF"))

#1 - censored & 2- progression
#1 - tumor_free & 2 with tumor .....neoplasm status 
#1 - NO & 2-YES .....TREATMENT CODE
#1 - no & 2-yes......radiation therapy
```

#https://rmvpaeme.github.io/KaplanMeier_intro/#exclude_censored_data:_yes_or_no 


\newpage
\section{subset neoplasm status of "With Tumor" and tumor_free}
```{r}
kdata<-newdata[newdata$Neoplasm_status=="with_tumor",]
k2data<-newdata[newdata$Neoplasm_status=="tumor_free",]

write.csv(kdata,"C://Users//Kelvin//Desktop//research with Dr. Leung//survival//neoplasm.status_withtumor.csv", row.names = FALSE)

write.csv(k2data,"C://Users//Kelvin//Desktop//research with Dr. Leung//survival//neoplasm.status_tumorfree.csv", row.names = FALSE)

#C://Users//Kelvin//Desktop//research with Dr. Leung//survival
```

\section{Alternate plot for uncensored cases}
```{r, warning=FALSE, message=FALSE}
library("survival")
library("survminer")

uncen_status<-newdata$progression_status=="2" #subset
fit2c<- survfit(Surv(newdata$progression_time, uncen_status) ~
                   newdata$Radiation.Therapy, data = newdata)
print(fit2c)

summary(fit2c)$table

ggsurvplot(fit2c,
          pval = TRUE, conf.int = F,
          #legend.labs=c("no_therapy", "therapy"),
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#E7B800", "#2E9FDF"))
```
