---
output: 
  pdf_document:
    keep_tex: true
    fig_caption: true
    latex_engine: xelatex
    number_sections: true
    toc: true
    toc_depth: 4
title: "SURVIVAL ANALYSIS - TCGA PRAD CANCER"
author: 
- Kelvin Ofori-Minta
- University of Texas at El Paso (UTEP) 
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontsize: 11pt
spacing: single
header-includes:
- \usepackage{amsmath}
- \usepackage{amssymb, bm}
- \usepackage{amsfonts}
- \usepackage{amsthm}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhf{}
- \rhead{Collaborative Research}
- \lhead{TCGA - Clinical Data}
- \cfoot{\thepage}
- \usepackage{algorithm}
- \usepackage[noend]{algpseudocode}
---

\newpage
\section{Loading and Cleaning Data}
```{r, warning=FALSE, message=FALSE}
data <- read.csv("neoplasm.status_tumorfree.csv", header = T, stringsAsFactors = F,
              na.strings = "NA")
#selecting columns of interest
#data<-s[, c(4,7,8,12,13,19,20,21,23,27:32,44:47,50,53:55,58,62,40:42)]
# write.csv(data,"C://Users//Kelvin//Desktop//Spring 2022//research with Dr. Leung//survival//selected_columns.csv")
```

\subsection{Inspecting dataframe for missing values}
```{r, warning=FALSE, message=FALSE}
require(inspectdf)
show_plot(inspect_na(data))
missing = inspect_na(data)
missing[ , 3] = round(missing[ ,3], 2)
names(missing) = c("variable", "count", "proportion")
require(kableExtra)
# missing<-as.matrix.data.frame(missing)
kable(missing)
# as.data.frame.matrix(missing)
# kable(as.da(missing))
```

\subsubsection{Rename long variables}
```{r}
"TMB-H means that the tumor has a high number of mutations. Doctors have found that
certain immunotherapy drugs are more likely to work
against TMB-H cancers. This is because the immune
system may be able to find and attack cancer cells with high
TMB more easily."


"Person neoplasm status...... You are correct, IMO:  tumor free does not mean normal, but rather whether (or not) the tumor (neoplasm) continues to be present.  It is a statement about the progression (or not) of the original disease; whereas normal is a statement that there was no disease to begin with."
```


\newpage
\subsubsection{Re-coding variables}
```{r, warning=FALSE, message=FALSE}

# newdata$Neodjuvant_Therapy_type <- factor(newdata$Neodjuvant_Therapy_type,
#                                     levels=c("No","Yes"),
#                                     labels=c("No","Yes")) all were "no"


data$In.PanCan.Pathway.Analysis<-factor(data$In.PanCan.Pathway.Analysis,
                                           levels=c("No","Yes"),
                                           labels=c("No","Yes"))


data$Prior.Diagnosis<-factor(data$Prior.Diagnosis, 
                                levels=c("No","Yes"),
                                labels=c("No","Yes"))
  

data$tissue_source.site<-factor(data$tissue_source.site,
                                   levels = c("university","Biotech & Pharma","Hospital","Research center"),
                                                          labels=c("university","biotech_pharma","hospital","research_centers"))


data$New_Tumor_event <- factor(data$New_Tumor_event,
                                    levels=c("No","Yes"),
                                    labels=c("No","Yes"))


data$Radiation.Therapy <- factor(data$Radiation.Therapy,
                                    levels=c("No","Yes"),
                                    labels=c("No","Yes"))

#all white , no adjuvant therapy
str(data)
```

\newpage
\section{KM Curve - survival probability with Radiation Therapy}
```{r, warning=FALSE, message=FALSE}
library("survival")
library("survminer")
ndata<-data
fit2a <- survfit(Surv(ndata$progression_time, ndata$progression_status) ~ ndata$Radiation.Therapy, data = ndata)
print(fit2a)

summary(fit2a)$table
```
```{r}
ggsurvplot(fit2a,
          #legend.labs=c("tumor_free", "with_tumor"),
          pval = TRUE, conf.int = F,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#E7B800", "#2E9FDF"))

#1 - censored & 2- progression
#1 - tumor_free & 2 with tumor .....neoplasm status 
#1 - NO & 2-YES .....TREATMENT CODE
```




\newpage
\section{KM Curve- survival probability:Censored cases of Radiation therapy}
```{r, warning=FALSE, message=FALSE}
library("survival")
library("survminer")

fit2b <- survfit(Surv(ndata$progression_time, ndata$progression_status==1) ~ ndata$Radiation.Therapy, 
               data = ndata)
print(fit2b)

summary(fit2b)$table
```
```{r}
ggsurvplot(fit2b,
          #legend.labs=c("tumor_free", "with_tumor"),
          pval = TRUE, conf.int = F,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#E7B800", "#2E9FDF"))

#1 - censored & 2- progression
#1 - tumor_free & 2 with tumor .....neoplasm status 
#1 - NO & 2-YES .....TREATMENT CODE
```


\newpage
\section{KM Curve- survival probability:Censored cases of New tumor event}
```{r, warning=FALSE, message=FALSE}
library("survival")
library("survminer")

fit2c <- survfit(Surv(ndata$progression_time, ndata$progression_status==2) ~ ndata$New_Tumor_event, 
               data = ndata)
print(fit2c)

summary(fit2c)$table
```
```{r}
ggsurvplot(fit2c,
          #legend.labs=c("tumor_free", "with_tumor"),
          pval = TRUE, conf.int = F,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#E7B800", "#2E9FDF"))

#1 - censored & 2- progression
#1 - tumor_free & 2 with tumor .....neoplasm status 
#1 - NO & 2-YES .....TREATMENT CODE
```


\newpage
\section{KM Curve- survival probability:Uncensored cases of Prior Diagnosis}
```{r, warning=FALSE, message=FALSE}
library("survival")
library("survminer")

fit2d <- survfit(Surv(ndata$progression_time, ndata$progression_status==2) ~ ndata$Prior.Diagnosis, 
               data = ndata)
print(fit2d)

summary(fit2d)$table
```
```{r}
ggsurvplot(fit2d,
          #legend.labs=c("tumor_free", "with_tumor"),
          pval = TRUE, conf.int = F,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#E7B800", "#2E9FDF"))

#1 - censored & 2- progression
```

